Concurrent Programming with CXL

### 2020-06-19 book/paper.tex
Massaging the text

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}
do not make a distinction between processes and threads.}) concurrently
\glsadd{concurrent execution}
read and update shared variables
\index{shared variable}
\glsadd{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
\glsadd{sequential execution}
program is mostly \emph{deterministic}.
\index{determinism}
\glsadd{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
\glsadd{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

Model checking is not a replacement for formal verification.
\index{formal verification}
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be relatively
simple.  In particular, it needs to have a relatively small number of
reachable states.
If the model is too large, it may run for longer than we may care to
wait for or even run out of memory.

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
\glsadd{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
\item Python is object-oriented; CXL is not;
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} shows a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

In CXL, constants have a default, but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ cxl -c N=100 triangle.cxl
#states = 103
no issues found
$
\end{verbatim}
\end{code}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
\glsadd{process}
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

The result in this particular interleaving is that \texttt{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}
\glsadd{race condition}
When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

\begin{enumerate}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \texttt{count}
was for each of the failed runs.
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{enumerate}

>>>>

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}
do not make a distinction between processes and threads.})
read and update shared variables
\index{shared variable}
concurrently and synchronize with one another makes programs more
complicated than programs where only one thing happens at a time.
\glsadd{concurrent execution}
\glsadd{shared variable}

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
\glsadd{sequential execution}
program is mostly \emph{deterministic}.
\index{determinism}
\glsadd{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity interfering with their execution.
\glsadd{atomicity}
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

Model checking is not a replacement for formal verification.
\index{formal verification}
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be
simple or it will take longer than we may care to wait for or run
out of memory.
In particular, the model needs to have a relatively small number of
reachable states.

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
\glsadd{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

\begin{itemize}
\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
CXL does support \texttt{for} loops and various comprehensions over sets.
\item Python is object-oriented, supporting classes with methods and
inheritance; CXL has objects but does not support classes.  CXL supports
pointers to objects.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} shows a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The example is sequential and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

In CXL, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ cxl -c N=100 triangle.cxl
#states = 103
no issues found
$
\end{verbatim}
\end{code}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
running in parallel while sharing variables.
\glsadd{process}
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits until \texttt{done} flags are set to \texttt{True}.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

The result in this particular interleaving is that \texttt{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}
When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.
\glsadd{race condition}

@@

\begin{enumerate}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \texttt{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{enumerate}
